---
title: "JMazon_Assignment3"
author: "John Mazon"
date: "2/14/2021"
output: 
  html_document:
    fig_height: 3
    fig_width: 5
  pdf_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
---

```{r, setup, include=FALSE}
require(tidyverse)
```

## Problem Set #1
### Question #1

What is the rank of matrix A where:

$$A = \left( \begin{array}{cccc}
1 & 2 & 3 & 4 \\
-1 & 0 & 1 & 3 \\
0 & 1 & -2 & 1\\
5 & 4 & -2 & -3 \end{array}\right)$$

To answer this, we perform gaussian elimination.  When in upper-diagonal form, the number of non-zero rows is the rank.

I first target $A_{21}$ and $A_{41}$:

$$\left( \begin{array}{cccc}
1 & 2 & 3 & 4 \\
-1 & 0 & 1 & 3 \\
0 & 1 & -2 & 1\\
5 & 4 & -2 & -3 \end{array}\right) \rightarrow
\left( \begin{array}{cccc}
1 & 2 & 3 & 4 \\
0 & 2 & 4 & 7 \\
0 & 1 & -2 & 1 \\
0 & -6 & -17 & -22 \end{array}\right)$$

Now $A_{32}$ and $A_{42}$:

$$\left( \begin{array}{cccc}
1 & 2 & 3 & 4 \\
0 & 2 & 4 & 7 \\
0 & 1 & -2 & 1 \\
0 & -6 & -17 & -22 \end{array}\right) \rightarrow
\left( \begin{array}{cccc}
1 & 2 & 3 & 4 \\
0 & 2 & 4 & 7 \\
0 & 0 & 8 & 5\\
0 & 0 & -5 & -1 \end{array}\right)$$

And lastly $A_{43}$:

$$\left( \begin{array}{cccc}
1 & 2 & 3 & 4 \\
0 & 2 & 4 & 7 \\
0 & 0 & 8 & 5\\
0 & 0 & -5 & -1 \end{array}\right) \rightarrow
\left( \begin{array}{cccc}
1 & 2 & 3 & 4 \\
0 & 2 & 4 & 7 \\
0 & 0 & 8 & 5\\
0 & 0 & 0 & 17 \end{array}\right)$$

Because no row is all zeroes, the rank of $A$ is 4.

As a check using R, we can compute the eigenvalues.  The number of non-zero eigenvalues is the rank of A.

```{r}
eigenvalues <- eigen(matrix(c(1, -1, 0, 5, 2, 0, 1, 4, 3, 1, -2, -2, 4, 3, 1, -3), nrow = 4))$values

print(paste("Number of non-zero eigenvalues (= rank(A)):", sum(abs(eigenvalues) > 1e-12)))
```

### Question #2
The maximum rank of a $m$ by $n$ matrix is equal to the smaller of $m$ and $n$.  This is because $p$-dimensional space is spanned by $p$ independent vectors.  Any additional vectors would be linearly dependent and matrix rank is equal to the number of linearly independent row (or column) vectors.

### Question #3
The rank of matrix $B$ is 1.  This is because rows 2 and 3 are multiples of row 1. 

## Problem Set #2
### Question #1

The characteristic equation is given by:

$$det\left(\mathbf{A}-\lambda \mathbf{I}\right) = 0$$

which for matrix A is:

$$\left | \begin{array}{ccc}
1 - \lambda & 2 & 3 \\
0 & 4 - \lambda & 5 \\
0 & 0 & 6 - \lambda \end{array}\right | = (1-\lambda)(4-\lambda)(6-\lambda) = 0$$

This implies that the eigenvalues of $A$ are 1, 4 and 6--the roots of the characteristic equation.

To compute the associated eigenvalues, we use the eigenvalue/vector equation:

$$\mathbf{Ax}=\lambda\mathbf{x}$$

For $\lambda = 1$ we have:

$$\left ( \begin{array}{ccc}
1  & 2 & 3 \\
0 & 4 & 5 \\
0 & 0 & 6 \end{array}\right )
\left ( \begin{array}{c}
x_1 \\
x_2 \\
x_3 \\ \end{array} \right ) = \left ( \begin{array}{c}
x_1 \\
x_2 \\
x_3 \\ \end{array} \right )$$

Because $A$ is already in reduced (row-echelon) form, we can solve immediately.  The last row implies that $x_3 = 0$.  Back-substituting into row 2 gives $x_2 = 0$.  Lastly, back-substituting into row 1 gives $x_1 = 1$, hence the first eigenvalue/vector pair is:

$$\lambda = 1\text{ and }x= \left ( \begin{array}{c}
1 \\
0 \\
0 \\ \end{array} \right )$$

Likewise, when $\lambda = 4$, we find that:

$$\left ( \begin{array}{ccc}
1  & 2 & 3 \\
0 & 4 & 5 \\
0 & 0 & 6 \end{array}\right )
\left ( \begin{array}{c}
x_1 \\
x_2 \\
x_3 \\ \end{array} \right ) = \left ( \begin{array}{c}
4x_1 \\
4x_2 \\
4x_3 \\ \end{array} \right )$$

Row 3 implies that $x_3$ equals 0.  Row 2 this implies that $x_2$ is non-zero (choosing $x_2 = 0$ leads to a trivial result).  Lastly, row 1 implies:

$$x_1 + 2x_2 = 4x_1$$
$$2x_2 = 3x_1$$
$$x_2 = \frac{3}{2}x_1$$

Or that the second eigenvector (normalized to unit length) is:

$$\lambda = 4\text{ and }x= \left ( \begin{array}{c}
0.555 \\
0.832 \\
0 \\ \end{array} \right )$$

Lastly, for $\lambda = 6$, we have:

$$\left ( \begin{array}{ccc}
1  & 2 & 3 \\
0 & 4 & 5 \\
0 & 0 & 6 \end{array}\right )
\left ( \begin{array}{c}
x_1 \\
x_2 \\
x_3 \\ \end{array} \right ) = \left ( \begin{array}{c}
6x_1 \\
6x_2 \\
6x_3 \\ \end{array} \right )$$

Row 3 tells us nothing about $x_3$, but choosing $x_3 = 0$ leads to a trivial result.  So we will let $x_3 = 1$.

Then row 2 implies:

$$4x_2 + 5x_3 = 6x_2$$
$$-2x_2=-5$$
$$x_2 = 5/2$$

Plugging these results into row 1 gives:

$$x_1 + 5 + 3 = 6x_1$$
$$5x_1=8$$
$$x_1 = 8/5$$
Thus the final eigenvalue/vector pair is:

$$\lambda = 6\text{ and }x= \left ( \begin{array}{c}
8/5 \\
5/2 \\
1 \\ \end{array} \right )$$

Normalizing $\mathbf{x}$ to unit length, we find:

$$\lambda = 6\text{ and }x= \left ( \begin{array}{c}
0.511 \\
0.798 \\
0.319 \\ \end{array} \right )$$

As a quick check using R:

```{r}
eigen(matrix(c(1,0,0,2,4,0,3,5,6), nrow = 3))
```

