---
title: "Assignment 9 - DATA 605"
author: "John Mazon"
output: 
  pdf_document:
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
---

```{r, setup, include=FALSE}
require(tidyverse)
```

### Q1 Pilsdorf Beer Company Stock Price

Let the stock price on day $n$ of the year be $Y_n$.  The differences, $X_n = Y_{n+1}-Y_n$ are observed to be independent and normally distributed $N(\mu = 0, \sigma^2 = 1/4)$.  If $Y_1=100$, predict the likelihood of $Y_{365} \ge 100$.  Repeat for values of 110 and 120.

I begin by noting that $X_2+X_1=Y_3-Y_1$.  By extension:

$$X_{364}+X_{363}+\cdots+X_1=Y_{365}-Y_1=Y_{365}-100$$

Thus, $Y_{365}$ can be expressed as:

$$Y_{365} = X_{364}+X_{363}+\cdots+X_1+100$$

The expected value of $Y_{365}$ follows from:

$$\mathbb{E}(Y_{365}) = \mathbb{E}\left[\sum_{i=1}^{364}X_i\right] + 100$$
$$\mathbb{E}(Y_{365}) = \sum_{i=1}^{364}\mathbb{E}\left[X_i\right] + 100$$

$$\mathbb{E}(Y_{365}) = 100$$
Because the $X_i$ are independent, the variance of $Y_{365}$ is the sum of variances of the $X_i$:

$$\mathbb{E}\left[(Y_{365}-100)^2\right]=\frac{364}{4}=91$$
Hence $Y_{365}$ is normally distributed with $\mu=100$ and $\sigma^2=91$.

By definition, $P\left(Y_{365}\ge100\right)=0.5$.  For values of 110 and 120, we use R's pnorm function.



```{r}
pge110 <- round(1 - pnorm(110, 100, 91), 4)
pge120 <- round(1 - pnorm(120, 100, 91), 4)
```

We find $P\left(Y_{365}\ge110\right)=$ `r pge110` and $P\left(Y_{365}\ge120\right)=$ `r pge120`.

## Q2 
Find the mean and variance of the binomial distribution from it's moment-generating function.

The MGF for the binomial distribution is:

$$M(t)=(q+pe^t)^n$$

To get the mean and variance, we compute the first two derivatives of $M(t)$:

$$M'(t)=npe^t(q+pe^t)^{n-1}$$
$$M''(t)=n(n-1)p^2e^{2t}(q+pe^t)^{n-2}+npe^t(q+pe^t)^{n-1}$$
The mean (first non-centered moment) is obtained by evaluating $M'(0)$:

$$\mu=M'(0)=np(q+p)^{n-1}=np(1-p+p)^{n-1}=np$$

The variance is obtained from:

$$\sigma^2=M''(0)-\mu^2=n(n-1)p^2+np-n^2p^2$$
$$\sigma^2=-np^2+np=np(1-p)=npq$$

## Q3 

Find the mean and variance of the exponential distribution from it's moment-generating function.

The MGF for the exponential distribution is:

$$M(t)=\frac{\lambda}{\lambda-t}\text{ for }t<\lambda$$
Taking the first and second derivatives yields:

$$M'(t)=\frac{\lambda}{(\lambda-t)^2}$$
$$M''(t)=\frac{2\lambda}{(\lambda-t)^3}$$
Thus the mean is:

$$\mu=M'(0)=\frac{1}{\lambda}$$
and the variance:

$$\sigma^2=M''(0)-\mu^2=\frac{2\lambda}{(\lambda)^3}-\frac{1}{\lambda}=\frac{1}{\lambda^2}$$
$$\sigma^2=\frac{1}{\lambda^2}$$







