---
title: 'DATA 605 - Homework #11'
author: "John Mazon"
date: "4/18/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Objective

Using the cars dataset in R, build a linear model for stopping distance as a function of speed and replicate the analysis of your textbook chapter 3 (visualization, quality evaluation of the model, and residual analysis).
The data give the speed of cars and the distances taken to stop. Note that the data were recorded in the 1920s. 

A data frame with 50 observations on 2 variables.
[,1] 	speed 	numeric 	Speed (mph)
[,2] 	dist 	numeric 	Stopping distance (ft)

### Analysis

```{r}
summary(cars)


```

Here we see the plot of two variables. We witness that stopping distance is the response var and Speed is the explanatory var

```{r}
plot(cars$speed, cars$dist, xlab='Speed (mph)', ylab='Stopping Distance (ft)', 
     main='Stopping Distance vs. Speed')

## here we will build the linear model and continue to find the best fitting line.

cars_lm <- lm(cars$dist ~ cars$speed)
cars_lm


plot(cars$speed, cars$dist, xlab='Speed (mph)', ylab='Stopping Distance (ft)', 
     main='Stopping Distance vs. Speed')
abline(cars_lm)

##we witness a correlation between two variables, but now we must evaluate the linear model

summary(cars_lm)
```


We see that in fact median value of the residuals is very close to zero and Quarts. and min/max values are about the same. 
Now it's evident that standard error of the speed variable is more than 9 times smaller than the corresponding coefficient.
We assume not a lot of variability in this coefficient. 

Seeing the other side, the difference between the intercept estimate and standard error is less significant. This means there may be more variability. 
The speed coefficient is highly significant. The intercept coefficient is less significant, but it is still relevant depending on the confidence interval desired.
At last we see, R2 explains about 65.11% of the data’s variation.

```{r}
plot(cars_lm$fitted.values, cars_lm$residuals, xlab='Fitted Values', ylab='Residuals')
abline(0,0)
```

We can say that the outlier values do not show the same variance of the residuals; we are assuming similar variance of residuals.

```{r}
qqnorm(cars_lm$residuals)
qqline(cars_lm$residuals)
```


the normal Q-Q plot of the residuals appears to follow the theoretical line. Residuals are distributed normally.


### Conclusion

We see the LM depicts are clear understanding of this data. There appears to be small curvature in the main plot and in the residuals plot

```{r}
speed <- cars$speed
speed2 <- speed^2
dist <- cars$dist
  
cars_qm <- lm(dist ~ speed + speed2)
summary(cars_qm)
```

```{r}
speedvalues <- seq(0, 25, 0.1)
predictedcounts <- predict(cars_qm,list(speed=speedvalues, speed2=speedvalues^2))

plot(speed, dist, pch=16, xlab='Speed (mph)', ylab='Stopping Distance (ft)')
lines(speedvalues, predictedcounts)
```


```{r}
plot(cars_qm$fitted.values, cars_qm$residuals, xlab='Fitted Values', ylab='Residuals')
abline(0,0)

qqnorm(cars_qm$residuals)
qqline(cars_qm$residuals)

```

